---
title: "Reproducible Research course Project 1"
author: "Mitali"
date: "6/28/2020"
output: html_document
---

## Loading and Preprocessing the data
```{r, echo = TRUE}
library(dplyr)
library(ggplot2)
setwd("/Users/mbarhanp/Documents/Coursera coursework/Reproducible Research")
activity <- read.csv("activity.csv")
```

Exploring the data
```{r}
dim(activity)
names(activity)
head(activity)
str(activity)
#total number of missing data
sum(is.na(activity$steps))/dim(activity)[[1]]
```

## Histogram of Total Number of Steps taken per day
**1. Number of steps per day**
```{r, echo = TRUE}
StepsPerDay <- aggregate(activity$steps, list(activity$date), FUN=sum)
colnames(StepsPerDay) <- c("Date", "Steps")
```
**2. Histogram of the total number of steps taken each day**
```{r histogram1, echo=TRUE, fig.width=10, warning=FALSE}
# draw the histogram
g <- ggplot(StepsPerDay, aes(Steps))
g+geom_histogram(boundary=0, binwidth=2500, col="darkgreen", fill="lightgreen")+ggtitle("Histogram of steps per day")+xlab("Steps")+ylab("Frequency")+theme(plot.title = element_text(face="bold", size=12))+scale_x_continuous(breaks=seq(0,25000,2500))+scale_y_continuous(breaks=seq(0,18,2))
```

## Mean and median number of steps taken per day
```{r echo=TRUE}
# Mean
mean(StepsPerDay$Steps, na.rm=TRUE)
#Median
median(StepsPerDay$Steps, na.rm=TRUE)
```

## Time series plot of the average number of steps taken
```{r timeplot1, echo=TRUE, fig.width=10, warning=FALSE}
# create table with steps per time
StepsPerTime <- aggregate(steps~interval,data=activity,FUN=mean,na.action=na.omit)
# variable time (more comprensible for the graph axis)
StepsPerTime$time <- StepsPerTime$interval/100
# draw the line plot
h <- ggplot(StepsPerTime, aes(time, steps))
h+geom_line(col="brown")+ggtitle("Average steps per time interval")+xlab("Time")+ylab("Steps")+theme(plot.title = element_text(face="bold", size=12))
```

## The 5-minute interval that, on average, contains the maximum number of steps
```{r echo=TRUE, fig.width=10, warning=FALSE}
# table for dplyr
ST <- tbl_df(StepsPerTime)
# find the column
ST %>% select(time, steps) %>% filter(steps==max(ST$steps))
```

## Code to describe and show a strategy for imputing missing data
**1. Total number of missing values in the dataset**

```{r echo=TRUE}
# table for dplyr
ACT <- tbl_df(activity)
# find the column
ACT %>% filter(is.na(steps)) %>% summarize(missing_values = n())
```

**2. Replace missing values**  
The rounded values of the average 5-minute interval is used to replace the NA values.  
*CompleteSteps* is the new column without missing values.

```{r echo=TRUE}
# values without NA are imputed in a new column
activity$CompleteSteps <- ifelse(is.na(activity$steps), round(StepsPerTime$steps[match(activity$interval, StepsPerTime$interval)],0), activity$steps)
```

**3. New dataset that is equal to the original dataset but with the missing data filled in**   
The first ten values of the new dataset are shown below.  

```{r echo=TRUE}
# new dataset activityFull
activityFull <- data.frame(steps=activity$CompleteSteps, interval=activity$interval, date=activity$date)
# see first 10 values of the new dataset
head(activityFull, n=10)
```

## Histogram of the total number of steps taken each day after missing values are imputed
```{r histogram2, echo=TRUE, fig.width=10, warning=FALSE}
# prepare data
StepsPerDayFull <- aggregate(activityFull$steps, list(activityFull$date), FUN=sum)
colnames(StepsPerDayFull) <- c("Date", "Steps")
# draw the histogram
g <- ggplot(StepsPerDayFull, aes(Steps))
g+geom_histogram(boundary=0, binwidth=2500, col="darkblue", fill="lightblue")+ggtitle("Histogram of steps per day")+xlab("Steps")+ylab("Frequency")+theme(plot.title = element_text(face="bold", size=12))+scale_x_continuous(breaks=seq(0,25000,2500))+scale_y_continuous(breaks=seq(0,26,2))
```

## Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
**1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**  
*DayType* is the new column indicating if the day is a weekday day or a weekend day: the first ten values of the new table are shown below

```{r echo=TRUE}
# Create variable with date in correct format
activityFull$RealDate <- as.Date(activityFull$date, format = "%Y-%m-%d")
# create a variable with weekdays name
activityFull$weekday <- weekdays(activityFull$RealDate)
# create a new variable indicating weekday or weekend
activityFull$DayType <- ifelse(activityFull$weekday=='Saturday' | activityFull$weekday=='Sunday', 'weekend','weekday')
# see first 10 values
head(activityFull, n=10)
```

**2. Two time series plot of the 5-minute interval (x) and the average number of steps taken averaged across weekday days or weekend days (y).**  

```{r timeplot2, echo=TRUE, fig.width=10, warning=FALSE}
# create table with steps per time across weekdaydays or weekend days
StepsPerTimeDT <- aggregate(steps~interval+DayType,data=activityFull,FUN=mean,na.action=na.omit)
# variable time (more comprensible for the graph axis)
StepsPerTimeDT$time <- StepsPerTime$interval/100
# draw the line plot
j <- ggplot(StepsPerTimeDT, aes(time, steps))
j+geom_line(col="darkred")+ggtitle("Average steps per time interval: weekdays vs. weekends")+xlab("Time")+ylab("Steps")+theme(plot.title = element_text(face="bold", size=12))+facet_grid(DayType ~ .)
```

